---
title: "Negative Binomial Regression"
author: 'Samuel Peterson, Elijah Wagner, and Ethan Olcott'
date: "April 17, 2024"
output:
  html_document:
    df_print: paged
header-includes: \usepackage{multirow, graphicx}
editor_options:
  chunk_output_type: console
urlcolor: blue
---

```{r setup, include = F}
library(NegativeBinomialRegression)
library(tidyverse)
library(GGally)
library(lubridate)
library(MASS)
<<<<<<< HEAD
library(dplyr)
library(AER)
library(MASS)
=======
library(AER)
library(MASS)
library(readxl)
>>>>>>> distribution1
```


## Negative Binomial Distribution

### 1 Brief Review

#### 1.1 The Binomial Distribution

The Binomial Distribution is a probability distribution that describes the number of successes in a fixed number of independent Bernoulli trials, where there is only success or failure as outcomes.

Some key assumptions and characteristics are:

1. Fixed Number of Trials
2. Independent Trials
3. Boolean Outcomes
4. Constant Probability
5. Discrete Distribution

#### 1.2 Connection to Negative Binomial Distribution

The Binomial Distribution predicts the number of successes when the number of trials is fixed. The Negative Binomial Distribution differs from this in that the number of successes is fixed, and the number of required trials is random.

### 2 The Negative Binomial

#### 2.1 Necessary Conditions

1. Independent Trials
2. Boolean Outcomes
3. Constant Probability
4. Trials Continue Till Desired Success Total Reached

The Negative Binomial Distribution has many of the same types of conditions as the Binomial Distribution (as to be expected), but the major difference comes in the fixed variable. The Negative Binomial is geared towards identifying the number of trials needed for a certain amount of trials.

#### 2.2 Math Behind Negative Binomial

The Probability Mass Function of the Negative Binomial Distribution with r = number of successes, p = probability of success, and k = the number of failures before the $r$th success is given by:

$\text{NegBin}(k\mid r, p) = \binom{k + r - 1}{r - 1} \cdot p^{r} \cdot (1 - p)^k\qquad\text{for}\qquad k,r=1, 2,\cdots\quad\text{and}\quad0<p\leq1$

The Expected Value and Variance of the Negative Binomial Distribution described above are as follows:

**Expected Value**: $\quad E(\mathbf X)=\displaystyle\frac{r\ \cdot\ (1 - p)}{p}$

**Variance**: $\quad V(\mathbf X)=\displaystyle\frac{r\ \cdot\ (1 - p)}{p^2}$

*Note: There are many variations of the Negative Binomial Distribution when using separate parameters. There are variations when you can count the number of trials, redefine the r as the number of failures, or other variations of them. These do change the PMF and $E(\mathbf X)$ and $V(\mathbf X)$ functions for the different variations.*

#### 2.3 Statistical Inference

Minimum Variance Unbiased Estimator for p:

If the probability of success isn't known and sampling will continue until the $r$th success is found, a sufficient statistic for the experiment then is $k$, the number of failures. In this case to estimate $p$, the minimum variance unbiased estimator is:

$$\hat{p} = \frac{r - 1}{r + k - 1}$$

Maximum Likelihood Estimate of $p$:

$$\hat{p} = \frac{r}{r + k}$$

This estimate is biased, however its inverse $\frac{r + k}{r}$ is an unbiased estimate of $\frac{1}{p}$

<<<<<<< HEAD

#### 2.4 Special Parametrization

Now see that if we re-parametrize with $\mu=\frac{r(1-p)}p$, $\alpha=\frac1r$, and $y=k$ that we get 
$$\text{NegBin}(y\mid\mu,\alpha)=\left(\begin{matrix}y+\frac1\alpha\\ \frac1\alpha-1\end{matrix}\right)\left(\frac1{1+\alpha\mu}\right)^{\frac1\alpha}\left(\frac{\alpha\mu}{1+\alpha\mu}\right)^y$$

## Negative Binomial Regression

Negative Binomial regression is very similar to Poisson regression and is applied when the data is a discrete, quantitative response variable ranging from 0 to infinity and larger counts are rare. The difference comes in with the dispersion. Since the Negative Binomial distribution is an extension of the Poisson distribution with the extra parameter $r$ the variance of the data does not have to be the same as the mean or expected value.

We therefore have the same link function for both Poisson and Negative Binomial regression, $\log(\widehat{E[Y\mid\mathrm x_i]})=\mathrm x_i\beta_i$ where $\mathrm x_i$ is the $i$th row of the model matrix $\mathbf X$.

Or equivalently since $E(Y)=\frac{r\ \cdot\ (1-p)}p=\mu$ we have $\log\left(\widehat{\mu\mid \mathrm x_i}\right)=\beta_0+\beta_1\mathrm x_i+\cdots+\beta_n\mathrm x_n=\mathbf X\boldsymbol\beta$

Therefore we have

$$\widehat{\mu\mid\mathrm x_i}=e^{\mathbf X\boldsymbol\beta}$$

Thus

$$p(y_i)=\left(\begin{matrix}y_i+1/\alpha\\1/\alpha-1\end{matrix}\right)\left(\frac{1}{1+\alpha e^{\mathbf X\boldsymbol\beta}}\right)^{1/\alpha}\left(\frac{\alpha e^{\mathbf X\boldsymbol\beta}}{1+\alpha e^{\mathbf X\boldsymbol\beta}}\right)^{y_i}$$

2.4 Connection to the Poisson Distribution
=======
#### 2.4 Connection to the Poisson Distribution
>>>>>>> distribution1

The Poisson Distribution and the Negative Binomial Distribution are both distributions that can be seen as similar in nature. It is possible to look at the Negative Binomial Distribution as the generalization of the Poisson Distribution. This can be done because the Negative Binomial Distribution converges to the Poisson Distribution as the number of trials goes to infinity and the probability of success is relatively small. The math of this is as follows:

\begin{align}
\lim_{r\to\infty}\text{NegBin}\Big(r,\frac{r}{r+\lambda}\Big)
&=\lim_{r\to\infty}\left(\frac{(k+r)!}{k!r!}\left(1-\frac{r}{r+\lambda}\right)^k\left(\frac{r}{r+\lambda}\right)^r\right)\\
&=\lim_{r\to\infty}\left(\frac{\lambda^k}{k!}\cdot\frac{(r+k)!}{r!(r+\lambda)^k}\cdot\frac{1}{\left(1+\frac{\lambda}{r}\right)^r}\right)\\
&=\frac{\lambda^k}{k!}\cdot1\cdot\frac{1}{e^\lambda}\\
&=\frac{\lambda^ke^{-\lambda}}{k!}\\
&=\text{Poisson}(\lambda)\qquad\blacksquare
\end{align}

<br>
Thus when $p=\dfrac{r}{r+\lambda}\Rightarrow\lambda=\dfrac{(1-p)r}p$ and $r\to\infty$ we have $\text{NegBin}(r,p)=\text{Poisson}(\lambda)$.

The Negative Binomial Distribution allows for the Variance and Mean to vary (which the Poisson Distribution does not allow for) and thus can deal with some data that the Poisson cannot. This is especially useful when the Poisson Distribution has issues with data that is over-dispersed and provides an alternative to the Quasi-Poisson option we explored in this class.

A brief example of this can be shown using the data from our Homework 5. 

```{r}
shipacc <- ship_accidents %>% 
  filter(
    service != 0
  )

psn_mod <- glm(incidents ~ ., data = shipacc, family = poisson)

nb_mod <- glm.nb(incidents ~ ., data = shipacc)

psn_pearson <- residuals(psn_mod, type = 'pearson')
lin_preds <- predict(psn_mod, type = 'link')
df_psn <- data.frame(lin_preds = lin_preds,
                 pearson = psn_pearson, model = "Poisson")

nb_pearson <- residuals(nb_mod, type = 'pearson')
lin_preds <- predict(nb_mod, type = 'link')
df_nb <- data.frame(lin_preds = lin_preds,
                 pearson = nb_pearson, model = "Negative Binomial")

df_combined <- rbind(df_psn, df_nb)

ggplot(df_combined, aes(x = lin_preds, y = pearson, color = model)) +
  geom_point() +
  geom_hline(yintercept = 0) +
  labs(x = 'Linear Predictors', y = 'Pearson Residuals')

ggplot(df_combined, aes(sample = pearson, color = model)) +
  stat_qq() +
  stat_qq_line() +
  labs(title = "QQ Plot: Pearson Residuals")
```

These can be slightly difficult to interpret, so we will conduct a goodness of fit test.

```{r, include = F}
p_val_psn <- 1 - pchisq(psn_mod$deviance, df = psn_mod$df.residual)
p_val_nb <- 1 - pchisq(nb_mod$deviance, df = nb_mod$df.residual)

df_3_psn <- psn_mod$df.residual
df_3_nb <- nb_mod$df.residual

test_stat_psn <- qchisq(1 - p_val_psn, df_3_psn)
test_stat_nb <- qchisq(1 - p_val_nb, df_3_nb)
```

We have strong evidence that the natural Poisson model does not fit the data, however the Negative Binomial model performs much better here. The Poisson P-Value for the goodness of fit test is 0.0000002 and the P-Value for the Negative Binomial goodness of fit test is 0.01. We do have evidence that the Negative Binomial isn't a great fit for this data as well, but it performs considerably better than the Poisson model.

Plotting the predicted incidents on top of the actual incidents gives us a good idea of how the models match up!

```{r}
psn_predictions <- exp(predict(psn_mod, type = "link"))
nb_predictions <- exp(predict(nb_mod, type = "link"))

index <- 1:length(shipacc$incidents)

df1 <- data.frame(
  index = rep(index, 3),
  incidents = c(shipacc$incidents, psn_predictions, nb_predictions),
  model = c(rep("Actual", length(shipacc$incidents)), rep("psn_predictions", length(psn_predictions)), rep("nb_predictions", length(nb_predictions)))
)

ggplot(df1, aes(x = index, y = incidents, color = model)) +
  geom_point() +
  geom_line() +
  labs(x = "Index", y = "Incidents", color = "Model")
```

Let's consider anther data set! 

<<<<<<< HEAD
```{r}
bikes <- bikes_bridges
=======
```{r, eval = F}
bikes <- read_xlsx("bikesonbridges.xlsx")
>>>>>>> distribution1
psn_mod2 <- glm(Total ~ ., data = bikes, family = poisson)
nb_mod2 <- glm.nb(Total ~ ., data = bikes)


psn_pearson2 <- residuals(psn_mod2, type = 'pearson')
lin_preds2 <- predict(psn_mod2, type = 'link')
df_psn2 <- data.frame(lin_preds = lin_preds2,
                 pearson = psn_pearson2, model = "Poisson")

nb_pearson2 <- residuals(nb_mod2, type = 'pearson')
lin_preds2 <- predict(nb_mod2, type = 'link')
df_nb2 <- data.frame(lin_preds = lin_preds2,
                 pearson = nb_pearson2, model = "Negative Binomial")

df_combined2 <- rbind(df_psn2, df_nb2)

ggplot(df_combined2, aes(x = lin_preds, y = pearson, color = model)) +
  geom_point() +
  geom_hline(yintercept = 0) +
  labs(x = 'Linear Predictors', y = 'Pearson Residuals')

ggplot(df_combined2, aes(sample = pearson, color = model)) +
  stat_qq() +
  stat_qq_line() +
  labs(title = "QQ Plot: Pearson Residuals")
```

```{r, include = F, eval = F}
p_val_psn2 <- 1 - pchisq(psn_mod$deviance, df = psn_mod$df.residual)
p_val_nb2 <- 1 - pchisq(nb_mod$deviance, df = nb_mod$df.residual)

df_3_psn2 <- psn_mod$df.residual
df_3_nb2 <- nb_mod$df.residual

test_stat_psn2 <- qchisq(1 - p_val_psn2, df_3_psn2)
test_stat_nb2 <- qchisq(1 - p_val_nb2, df_3_nb2)
```

```{r, eval = F}
psn_predictions2 <- exp(predict(psn_mod2, type = "link"))
nb_predictions2 <- exp(predict(nb_mod2, type = "link"))

index <- 1:length(bikes$Total)

df1 <- data.frame(
  index = rep(index, 3),
  incidents = c(bikes$Total, psn_predictions2, nb_predictions2),
  model = c(rep("Actual", length(bikes$Total)), rep("psn_predictions2", length(psn_predictions2)), rep("nb_predictions2", length(nb_predictions2)))
)

ggplot(df1, aes(x = index, y = incidents, color = model)) +
  geom_point() +
  geom_line() +
  labs(x = "Index", y = "Incidents", color = "Model")
```

#### 2.5 Special Parametrization

Referencing what we had before once more:

\begin{align}
\lim_{r\to\infty}\text{NegBin}\Big(r,\frac{r}{r+\lambda}\Big)
&=\lim_{r\to\infty}\left(\frac{(k+r)!}{k!r!}\left(1-\frac{r}{r+\lambda}\right)^k\left(\frac{r}{r+\lambda}\right)^r\right)\\
&=\lim_{r\to\infty}\left(\frac{\lambda^k}{k!}\cdot\frac{(r+k)!}{r!(r+\lambda)^k}\cdot\frac{1}{\left(1+\frac{\lambda}{r}\right)^r}\right)\\
&=\frac{\lambda^k}{k!}\cdot1\cdot\frac{1}{e^\lambda}\\
&=\frac{\lambda^ke^{-\lambda}}{k!}\\
&=\text{Poisson}(\lambda)\qquad\blacksquare
\end{align}

Now see that if we re-parametrize with $\mu=\frac{r(1-p)}p$, $\alpha=\frac1r$, and $y=k$ that we get 
$$\text{NegBin}(y\mid\mu,\alpha)=\left(\begin{matrix}y+\frac1\alpha\\ \frac1\alpha-1\end{matrix}\right)\left(\frac1{1+\alpha\mu}\right)^{\frac1\alpha}\left(\frac{\alpha\mu}{1+\alpha\mu}\right)^y$$

## Negative Binomial Regression

Negative Binomial regression is very similar to Poisson regression and is applied when the data is a discrete, quantitative response variable ranging from 0 to infinity and larger counts are rare. The difference comes in with the dispersion. Since the Negative Binomial distribution is an extension of the Poisson distribution with the extra parameter $r$ the variance of the data does not have to be the same as the mean or expected value.

We therefore have the same link function for both Poisson and Negative Binomial regression, $\log(\widehat{E[Y\mid\mathrm x_i]})=\mathrm x_i\beta_i$ where $\mathrm x_i$ is the $i$th row of the model matrix $\mathbf X$.

Or equivalently since $E(Y)=\frac{r\ \cdot\ (1-p)}p=\mu$ we have $\log\left(\widehat{\mu\mid \mathrm x_i}\right)=\beta_0+\beta_1\mathrm x_i+\cdots+\beta_n\mathrm x_n=\mathbf X\boldsymbol\beta$

Therefore we have

$$\widehat{\mu\mid\mathrm x_i}=e^{\mathbf X\boldsymbol\beta}$$

Thus

$$p(y_i)=\left(\begin{matrix}y_i+1/\alpha\\1/\alpha-1\end{matrix}\right)\left(\frac{1}{1+\alpha e^{\mathbf X\boldsymbol\beta}}\right)^{1/\alpha}\left(\frac{\alpha e^{\mathbf X\boldsymbol\beta}}{1+\alpha e^{\mathbf X\boldsymbol\beta}}\right)^{y_i}$$


# Negative Binomial Regression


### Negative Binomial Regression Assumptions

1. **Negative Binomial Response**: The response variable is conditionally a count unit of time or space, described by the Negative Binomial Distribution. (Remember that the Poisson is a special limiting Negative Binomial result)
2. **Independence**: The observations are independent of one another.
3. **Linearity**: The log of the mean rate must be a linear function of $\mathrm x$.
4. **over-dispersed**: The variance is larger than the mean giving and overdispersion of the data.


### Parameter Estimation

The parameters $r, \beta_0, \beta_1, \cdots, \beta_p$ are also estimated using **maximum likelihood estimation** similar to Poisson regression. See that by using the special parametrization we have the following,

\begin{align}
L(\alpha,\beta)
&=\prod_{i=1}^np(y_i)\\
&=\prod_{i=1}^n\left(\begin{matrix}y_i+1/\alpha\\1/\alpha-1\end{matrix}\right)\left(\frac1{1+\alpha e^{\mathrm x_i\cdot\boldsymbol\beta}}\right)^{1/\alpha}\left(\frac{\alpha e^{\mathrm x_i\cdot\boldsymbol\beta}}{1+\alpha e^{\mathrm x_i\cdot\boldsymbol\beta}}\right)^{y_i}\\
&=\prod_{i=1}^n\frac{(y_i+1/\alpha)!}{(y_i+1)!(1/\alpha)!}\left(\frac1{1+\alpha e^{\mathrm x_i\cdot\boldsymbol\beta}}\right)^{1/\alpha}\left(\frac{\alpha e^{\mathrm x_i\cdot\boldsymbol\beta}}{1+\alpha e^{\mathrm x_i\cdot\boldsymbol\beta}}\right)^{y_i}\\
\end{align}

Thus we have log likelihood function,

\begin{align}
    \ell(\alpha,\boldsymbol\beta)
    &=\ln\left(L(\alpha,\boldsymbol\beta)\right)\\
    &=\ln\left(\prod_{i=1}^n\frac{(y_i+1/\alpha)!}{(y_i+1)!(1/\alpha)!}\left(\frac1{1+\alpha e^{\mathrm x_i\cdot\boldsymbol\beta}}\right)^{1/\alpha}\left(\frac{\alpha e^{\mathrm x_i\cdot\boldsymbol\beta}}{1+\alpha e^{\mathrm x_i\cdot\boldsymbol\beta}}\right)^{y_i}\right)\\
    &=\sum_{i=1}^n\Big[\ln\left((y_1+1/\alpha)!\right)-\ln\left((y_i+1)!\right)-\ln\left((1/\alpha)!\right)+y_i\ln(\alpha)+y_i\mathrm x_i\cdot\boldsymbol\beta-(y_i+1/\alpha)\ln\left(1+\alpha e^{\mathrm x_i\boldsymbol\beta}\right)\Big]
\end{align}

Then clearly the values of $\alpha$ and $\boldsymbol\beta$ that maximize $\ell(\alpha,\boldsymbol\beta)$ will be the maximum likelihood estimates and that we can solve for $\alpha$ and $\boldsymbol\beta$ by a standard derivative test, by use of the Hessian matrix, or by using a gradient descent method.



<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>

$$\mathbf Y_i\mid\boldsymbol\beta_0,\boldsymbol\beta_1,\cdots,\boldsymbol\beta_p,r\;\overset{\text{ind}}\sim\;\text{NegBin}(\mu_i,r)
\quad\text{with}\quad
\ln(\mu_i)=\boldsymbol\beta_0+\boldsymbol\beta_1\mathbf X_{i1}+\cdots+\boldsymbol\beta_p\mathbf X_{ip}$$

with priors:
\begin{align}
\boldsymbol\beta_0\;&\sim\;N(m_0,s_0^2)\\
\boldsymbol\beta_1\;&\sim\;N(m_1,s_1^2)\\
&\vdots\\
\boldsymbol\beta_p\;&\sim\;N(m_p,s_p^2)\\
r\;&\sim\;\text{Exp}(\cdots)
\end{align}


### Regression Requirements



### Examples


```{r, include = F}
droughts_nb <- droughts |>
  mutate(length = as.integer(length),
         year = as.character(year)) |>
  glm.nb(data = _, length ~ year)

ggplot(droughts, aes(x = length)) +
  geom_histogram(
    bins = length(unique(as.integer(droughts$length))), 
    fill = 'dodgerblue',
    color = 'black'
  )

predict(droughts_nb, data.frame(year = '1980'))
summary(droughts_nb)
plot(droughts_nb)

```

```{r, include = F}
bike_rentals$dteday <- as.Date(bike_rentals$dteday)
str(bike_rentals)

ggplot(bike_rentals, aes(x = cnt)) +
  geom_bar(color = 'dodgerblue')

mean(bike_rentals$cnt)
var(bike_rentals$cnt)
length(unique(bike_rentals$cnt))
```


```{r BikeRentalGGpairs, cache = T, fig.width = 15, fig.height = 15, include = F}
bike_rentals |>
  subset(select = -c(instant, dteday)) |>
  ggpairs()
```


### Outcomes and Comparisons